---
title: "Injury Analysis"
author: "Carmen Canedo"
date: "1 December 2020"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

# Loading libraries
```{r, message=FALSE, warning=FALSE}
source("setupCarmen.R")
```

# Hypothesis
The third hypothesis we are looking predicts that sports with high instances of injury will also have steady declines in participation.

# Data
The data used are as follows:

  + Youth emergency visits among US children for sports-related brain injuries from 2010-2016
  + US annual average emergency department visits for all sports injuries 2010-2016
  + US emergency department visits ages 5 to 24 from 2010 to 2016 by gender and activity
  + Percent of concussions in US youth that play sports as of 2017
  + Survey on the long-term effects of concussions in sports
  + Average annual number of US college female sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college male sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college for all seasons from 2009 to 2014 by gender
  + High School Participation Archive
  + Program for Injury Prevention, Education, & Research

# Injuries
## How many kids are getting hurt?
```{r Youth TBI Carmen}
# Getting rid of total counts
youth_ed_tbi_visits <- youth_ed_tbi_visits %>% 
  filter(Sport != "Total contact count") %>%
  filter(Sport != "Total limited contact count")

# Plotting bar graph
youth_ed_tbi_visits %>% 
  ggplot(aes(x = reorder(Sport, Count))) +
  geom_col(aes(y = Count,
               fill = `Contact Sport`)) +
  ylim(0, 55000) +
  scale_fill_manual(values = c("TRUE" = "#6eb4b8", "FALSE" = "#b8816e")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Youth Emergency Room Visits among US Children for \nSports-Related Brain Injuries",
       subtitle = "Years: 2010 - 2016",
       caption = "Centers for Disease Control and Prevention (CDC): \n Morbidity and Mortality Weekly Report (March 2019)",
       x = "Sport",
       y = "Count")
```

## What demographics are most affected?
```{r Concussions by Gender Carmen}
# Pivoting longer
num_concussions_2017 <- num_concussions_2017 %>% 
  pivot_longer(cols = c(male, female)) %>% 
  rename(gender = name)

# Plotting Number of Concussions by Gender
num_concussions_2017 %>% 
   ggplot(aes(x = concussions, y = value, fill = gender)) +
   geom_col(position = "dodge") +
  scale_fill_manual(values = c("female" = "#706eb8",
                               "male" = "#91b86e"),
                    name = "Gender") +
  labs(title = "Percent of Concussions in US Youth That Play Sports by Gender",
       subtitle = "Year: 2017",
       caption = "CDC: Self-Reported Concussions from Playing a Sport or Being Physically Active
Among High School Students",
       x = "",
       y = "Percent") +
  theme_minimal()
```

## What sports have the most injuries?
### Injuries by sport
```{r Combining Injury Totals Carmen}
# Iterating over and combining injury datasets
total_sport_injuries <- map_df(sports_injuries_type, combine_injury_totals)
```

```{r Lengthening Total Sports Injuries Carmen}
# Pivoting total sports injuries longer
total_sport_injuries <- Sport %>% 
  bind_cols(total_sport_injuries) %>% 
  select(-Activity) %>% 
  pivot_longer(cols = -Sport, names_to = "Injury Type", values_to = "Count")
```

```{r Boxplot of Injuries by Sport Carmen}
# Vector of colors
sports_colors <- c("#91b86e", "#956eb8", "#b86e90", "#6eb895", "#b8816e", "#b6b86e", "#706eb8")

# Plotting injuries by sport
total_injuries_boxplot <- total_sport_injuries %>% 
  ggplot(aes(x = Sport, y = Count, fill = Sport)) +
  geom_boxplot() +
  scale_fill_manual(values = sports_colors) +
  scale_y_continuous(breaks = seq(0, 175000, 25000)) +
  theme_minimal()
```

```{r Boxplot of Injuries by Sport Continued Carmen}
# Adding appropriate labels to histogram
total_injuries_boxplot +
  labs(title = "Variations in Amount of Injuries by Sport",
       subtitle = "Year: 2018",
       caption = "Colorado School of Public Health: \n Program for Injury Prevention, Education & Research") + 
  theme(axis.text.x = element_text(size = 7)) 
```

### Injuries by type
```{r Boxplot of Injuries by Type Carmen}
# Vector of colors
injury_colors <- c("#91b86e", "#956eb8", "#b86e90", "#6eb895", "#b8816e")

# Plotting boxplots of injury types
injury_type_boxplot <- total_sport_injuries %>% 
  ggplot(aes(x = `Injury Type`, y = Count, fill = `Injury Type`)) +
  geom_boxplot() +
  scale_fill_manual(values = injury_colors) +
  scale_y_continuous(breaks = seq(0, 175000, 25000)) +
  theme_minimal()

# Adding labels to graph
injury_type_boxplot +
  labs(title = "Variation in Most Common Injury Types",
       subtitle = "Year 2018",
       caption = "Colorado School of Public Health: \n Program for Injury Prevention, Education & Research")
```

## How have injuries rates changed over time?
### Diagnosis by year, color by type
```{r Lengthening Diagnosis Table Carmen}
# Pivoting diagnosis dataset
diagnosis_by_year <- diagnosis_by_year %>% 
  rename(Diagnosis = X1) %>% 
  filter(Diagnosis != "Total") %>% 
  pivot_longer(cols = !Diagnosis, names_to = "Year", values_to = "Percent")
```

```{r Formatting Diagnosis Carmen}
# Temporary variable holding all year values
temp_year <- diagnosis_by_year$Year

# Temporary variable holding all percent values
temp_perc <- diagnosis_by_year$Percent

# Using regex to clean observations
temp_year <- temp_year %>% 
  str_replace_all(pattern = ",", replacement = "") %>% 
  str_replace_all(pattern = "[[:punct:]]\\d{2}", replacement = "") %>% 
  as.integer()

# Using regex to get rid of percent sign
temp_perc <- temp_perc %>% 
  str_replace_all(pattern = "%", replacement = "") %>% 
  as.numeric()

# Joining new columns and renaming variables for clarity
diagnosis_by_year <- diagnosis_by_year %>% 
  select(-Year, -Percent) %>% 
  bind_cols(temp_year, temp_perc) %>% 
  rename(Year = `...2`) %>% 
  rename(Percent = `...3`)
```

```{r Scatterplot for Diagnosis by Sport Carmen}
# Creating scatter plot of diagnoses over time by type
diagnosis_time <- diagnosis_by_year %>% 
  ggplot(aes(x = Year, y = Percent, color = Diagnosis)) +
  facet_wrap(~ Diagnosis, scales = "free") +
  geom_point() +
  geom_smooth() +
  scale_color_manual(values = injury_colors) +
  theme_minimal()
```

```{r Scatterplot Formatting Carmen}
# Adding appropriate labels to diagram
diagnosis_time +
  labs(title = "Trends in Injury Diagnosis",
       subtitle = "Years: 2005 - 2018",
       caption = "Colorado School of Public Health: Program for Injury Prevention, Education & Research")
```


# Participation
## Who participates in sports?
```{r Map Data Carmen}
# Loading latitude and longitude data
usa_map <- map_data("state")
```

### Boys Participants Map
```{r Boy Participation Carmen}
# Lengthening boys participation data
boys_participation_long <- all_boys_athletes_states %>% 
  pivot_longer(cols = !State, names_to = "Sports", values_to = "Count")

# Making state names lowecase to match `usa_map`
boys_participation_long$State <- tolower(boys_participation_long$State)

# Plotting choropleth map
boys_participation_map <- boys_participation_long %>% 
  ggplot(aes(fill = Count)) +
  geom_map(aes(map_id = State), map = usa_map) +
  expand_limits(x = usa_map$long, y = usa_map$lat) +
  facet_wrap( ~ Sports)
```

```{r Boy Participation Map Carmen}
# Adding labels to map
boys_participation_map +
  theme_void() +
  labs(title = "Counts of Participation in Boy's Sports by State",
       subtitle = "Year: 2018",
       caption = "High School Participation Survey Archive") +
    theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_distiller(palette = "Greens", trans = "reverse")
```


### Boys Programs Availability Map
```{r Boys Programs Carmen}
# Lengthening program data
boys_programs_long <- all_boys_programs_states %>% 
  pivot_longer(cols = !State, names_to = "Sports", values_to = "Count")
  
# Making state names lowercase
boys_programs_long$State <- tolower(boys_programs_long$State)

# Plotting choropleth map
boys_programs_count <- boys_programs_long %>% 
  ggplot(aes(fill = Count)) +
  geom_map(aes(map_id = State), map = usa_map) +
  expand_limits(x = usa_map$long, y = usa_map$lat) +
  facet_wrap( ~ Sports)
```

```{r Boys Programs Map Carmen}
# Adding lables
boys_programs_count +
  theme_void() +
  labs(title = "Counts of Programs for Boy's Sports by State",
       subtitle = "Year: 2018",
       caption = "High School Participation Survey Archive") +
    theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_distiller(palette = "Purples", trans = "reverse")
```


### Girls Participants Map
```{r Gir Participation Carmen}
# Lengthening girls participation data
girls_participation_long <- girl_athletes_states %>% 
  select(-starts_with("X"), -Football) %>% 
  pivot_longer(cols = !State, names_to = "Sports", values_to = "Count")
  
# Making state names lowercase
girls_participation_long$State <- tolower(girls_participation_long$State)

# Plotting choropleth map
girls_participation_map <- girls_participation_long %>% 
  ggplot(aes(fill = Count)) +
  geom_map(aes(map_id = State), map = usa_map) +
  expand_limits(x = usa_map$long, y = usa_map$lat) +
  facet_wrap( ~ Sports)
```

```{r Girl Participation Map}
# Adding labels
girls_participation_map +
  theme_void() +
  labs(title = "Counts of Participation in Girl's Sports by State",
       subtitle = "Year: 2018",
       caption = "High School Participation Survey Archive") +
    theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_distiller(palette = "Greens", trans = "reverse")
```


### Girls Progams Map
```{r Girls Programs Carmen}
# Lengthening data
girls_programs_long <- girl_programs_states %>% 
  select(-starts_with("X"), -Football) %>% 
  pivot_longer(cols = !State, names_to = "Sports", values_to = "Count")

# Making state names lowercase
girls_programs_long$State <- tolower(girls_programs_long$State)

# Plotting chropleth map
girls_programs_count <- girls_programs_long %>% 
  ggplot(aes(fill = Count)) +
  geom_map(aes(map_id = State), map = usa_map) +
  expand_limits(x = usa_map$long, y = usa_map$lat) +
  facet_wrap( ~ Sports)
```

```{r Girl Participation Carmen}
# Adding labels
girls_programs_count +
  theme_void() +
  labs(title = "Counts of Programs for Girls's Sports by State",
       subtitle = "Year: 2018",
       caption = "High School Participation Survey Archive") +
    theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_distiller(palette = "Purples", trans = "reverse")
```

# Conclusion
## Public opinion
```{r Factor Levels Carmen}
# Vector of statements from survey
text_statements <- c("Players who suffer a head injury should be required to take a minimum set amount of time off from playing to recover.",
                     "Helmets should be changed to better protect against concussions.",
                     "The risks of playing football are widely known and players participate at their own risk.",
                     "There should be a standardized test used to determine if and when injured players at all levels may return to the field.",
                     "Aggressive tackles which are more prone to leading to head injuries should be restricted in youth football.")

# Vector of survey levels
survey_levels <- c(
  "Strongly agree",
  "Somewhat agree",
  "Somewhat disagree",
  "Strongly disagree",
  "Not at all sure"
) %>% 
  as_factor()
```

```{r Lengthening Data Carmen}
# Getting rid of Statement column to convert numbers to percents
survey_interim <- survey_tbi_effects %>% 
  select(-Statement)

# Calculating percentages
data_percentage <- survey_interim %>% 
  apply(2,
        function(x){
          perc <- x / sum(x)
          round(perc, digits = 2)
          })

# Converting to tibble and adding final column so it equals 1
diff_interim <- data_percentage %>% 
  as_tibble() %>% 
  select_if(is.numeric) %>% 
  mutate(sum = rowSums(.[1:5])) %>% 
  mutate(diff = 1 - sum)

# Saving into data_percentage
data_percentage <- diff_interim %>% 
  select(-sum)
  

# Adding the statement column back
data_percentage <- cbind(survey_tbi_effects$Statement, data_percentage) %>% 
  as_tibble()  %>% 
  slice(1:5)


# Renaming column
data_percentage <- data_percentage %>% 
  rename(Statement = `survey_tbi_effects$Statement`)

# Pivot long
long_percentage <- data_percentage %>% 
  pivot_longer(cols = contains("r")) %>% 
  rename(response = name)

```

```{r Stacked Bar Graph Carmen}
# Vector of colors and lavels
set_colors <- c("#91b86e", "#956eb8", "#b86e90", "#6eb895", "#b8816e", "#b6b86e")
xlabels <- c("1", "2", "3", "4", "5")

# Making graph
stacked_survey <- long_percentage %>% 
  ggplot(aes(x = Statement, y = value, fill = response)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = set_colors, name = "Response") +
  scale_x_discrete(labels = xlabels) +
  labs(title = "Survey Responses on The Long-Term Effects of Concussions in Sports",
       subtitle = "2014",
       caption = "Source: Harris Interactive | The Harris Poll #92",
       y = "") +
  theme_light()

stacked_survey
```
