---
title: "Injury Analysis"
author: "Carmen Canedo"
date: "11 October 2020"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Loading libraries
```{r}
library(tidyverse)
library(likert)
```

# Reading in data
The data used are as follows:

  + Youth emergency visits among US children for sports-related brain injuries from 2010-2016
  + US annual average emergency department visits for all sports injuries 2010-2016
  + US emergency department visits ages 5 to 24 from 2010 to 2016 by gender and activity
  + Percent of concussions in US youth that play sports as of 2017
  + Survey on the long-term effects of concussions in sports
  + Average annual number of US college female sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college male sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college for all seasons from 2009 to 2014 by gender
  
```{r}
youth_ed_tbi_visits <- read_csv("data/injury-data/ed-visits-among-us-children-for-sports-related-brain-injury-2010-2016-by-sport.csv")
general_ed_visits <- read_csv("data/injury-data/us-annual-average-emergency-department-visits-for-sports-injuries-2010-2016-by-age.csv")
sport_ed_visits <- read_csv("data/injury-data/us-emergency-department-visits-ages-5-24-2010-2016-by-gender-and-activity.csv")
num_concussions_2017 <- read_csv("data/injury-data/number-of-concussions-suffered-by-us-youth-that-played-sports-as-of-2017-by-gender.csv")
survey_tbi_effects <- read_csv("data/injury-data/survey-on-concussions-suffered-in-sports-and-long-term-brain-damage---effects-2014.csv")
female_college_sports <- read_csv("data/injury-data/annual-number-of-us-college-female-sports-injuries-during-fall-season-2009-2014.csv")
male_college_sports <- read_csv("data/injury-data/average-number-of-male-sports-injuries-during-us-college-fall-season-2009-2014.csv")
avg_college_sports <- read_csv("data/injury-data/annual-number-of-college-sports-injuries-in-the-us-by-gender-2009-2014.csv")
```

# Bar Graph: Youth emergency visits for brain injuries
```{r}
# Getting rid of total counts
youth_ed_tbi_visits <- youth_ed_tbi_visits %>% 
  filter(Sport != "Total contact count") %>%
  filter(Sport != "Total limited contact count")

# Plotting bar graph
youth_ed_tbi_visits %>% 
  ggplot(aes(x = reorder(Sport, Count))) +
  geom_col(aes(y = Count,
               fill = `Contact Sport`)) +
  ylim(0, 55000) +
  scale_fill_manual(values = c("TRUE" = "#6eb4b8", "FALSE" = "#b8816e")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Youth Emergency Room Visits among US Children for \nSports-Related Brain Injuries",
       subtitle = "Years: 2010 - 2016",
       caption = "Centers for Disease Control and Prevention (CDC): \n Morbidity and Mortality Weekly Report (March 2019)",
       x = "Sport",
       y = "Count")
```


# Histogram

# Count

# Bar Graph: Number of concussions by gender
```{r}
num_concussions_2017 <- num_concussions_2017 %>% 
  pivot_longer(cols = c(male, female)) %>% 
  rename(gender = name)


num_concussions_2017 %>% 
   ggplot(aes(x = concussions, y = value, fill = gender)) +
   geom_col(position = "dodge") +
  scale_fill_manual(values = c("female" = "#706eb8",
                               "male" = "#91b86e")) +
  labs(title = "Percent of Concussions in US Youth That Play Sports by Gender",
       subtitle = "Year: 2017",
       caption = "CDC: Self-Reported Concussions from Playing a Sport or Being Physically Active
Among High School Students",
       x = "",
       y = "Percent") +
  theme_minimal()
```

# Stacked Bar Graph
```{r}
survey_levels <- c("Not at all sure",
                   "Strongly disagree",
                   "Somewhat disagree",
                   "Somewhat agree",
                   "Strongly agree")

# Getting rid of Statement column to convert numbers to percents
survey_interim <- survey_tbi_effects %>% 
  select(-Statement)

# Calculating percentages
data_percentage <- survey_interim %>% 
  apply(2,
        function(x){
          perc <- x / sum(x)
          round(perc, digits = 2)
          })

# Adding the statement column back
data_percentage <- cbind(survey_tbi_effects$Statement, data_percentage) %>% 
  as_tibble() %>% 
  group_by(-V1) %>% 
  summarise(diff = 1 - sum())

# Renaming column
data_percentage <- data_percentage %>% 
  rename(statement = V1)

# Pivot long
long_percentage <- data_percentage %>% 
  pivot_longer(cols = contains("r")) %>% 
  rename(response = name)

```

```{r}
set_colors <- c("#91b86e", "#956eb8", "#b86e90", "#6eb895", "#b8816e", "#b6b86e")

# Making graph
data_percentage %>% 
  ggplot(aes(y = value, x = statement)) +
  geom_col(position = "fill")
```

