---
title: "Injury Analysis"
author: "Carmen Canedo"
date: "11 October 2020"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

# Loading libraries
```{r, message=FALSE, warning=FALSE}
source("setupCarmen.R")
```

# Hypothesis
The third hypothesis we are looking predicts that sports with high instances of injury will also have steady declines in participation.

# Data
The data used are as follows:

  + Youth emergency visits among US children for sports-related brain injuries from 2010-2016
  + US annual average emergency department visits for all sports injuries 2010-2016
  + US emergency department visits ages 5 to 24 from 2010 to 2016 by gender and activity
  + Percent of concussions in US youth that play sports as of 2017
  + Survey on the long-term effects of concussions in sports
  + Average annual number of US college female sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college male sports injuries during the fall season from 2009 to 2014
  + Average annual number of US college for all seasons from 2009 to 2014 by gender
  + High School Participation Archive
  + Program for Injury Prevention, Education, & Research

# Injuries
## How many kids are getting hurt?
Bar Graph: Youth emergency visits for brain injuries
```{r}
# Getting rid of total counts
youth_ed_tbi_visits <- youth_ed_tbi_visits %>% 
  filter(Sport != "Total contact count") %>%
  filter(Sport != "Total limited contact count")

# Plotting bar graph
youth_ed_tbi_visits %>% 
  ggplot(aes(x = reorder(Sport, Count))) +
  geom_col(aes(y = Count,
               fill = `Contact Sport`)) +
  ylim(0, 55000) +
  scale_fill_manual(values = c("TRUE" = "#6eb4b8", "FALSE" = "#b8816e")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Youth Emergency Room Visits among US Children for \nSports-Related Brain Injuries",
       subtitle = "Years: 2010 - 2016",
       caption = "Centers for Disease Control and Prevention (CDC): \n Morbidity and Mortality Weekly Report (March 2019)",
       x = "Sport",
       y = "Count")
```

## What demographics are most affected?
Bar Graph: Number of concussions by gender
```{r}
num_concussions_2017 <- num_concussions_2017 %>% 
  pivot_longer(cols = c(male, female)) %>% 
  rename(gender = name)


num_concussions_2017 %>% 
   ggplot(aes(x = concussions, y = value, fill = gender)) +
   geom_col(position = "dodge") +
  scale_fill_manual(values = c("female" = "#706eb8",
                               "male" = "#91b86e"),
                    name = "Gender") +
  labs(title = "Percent of Concussions in US Youth That Play Sports by Gender",
       subtitle = "Year: 2017",
       caption = "CDC: Self-Reported Concussions from Playing a Sport or Being Physically Active
Among High School Students",
       x = "",
       y = "Percent") +
  theme_minimal()
```

## What sports have the most injuries?
### Boxplots of injuries by sports
```{r}
total_sport_injuries <- map_df(sports_injuries_type, combine_injury_totals)
```

```{r}
total_sport_injuries <- Sport %>% 
  bind_cols(total_sport_injuries) %>% 
  select(-Activity) %>% 
  pivot_longer(cols = -Sport, names_to = "Injury Type", values_to = "Count")
```

```{r}
total_sport_injuries %>% 
  ggplot(aes(x = Sport, y = Count)) +
  geom_boxplot()
```
### Boxplot of injuries by type
```{r}
total_sport_injuries %>% 
  ggplot(aes(x = `Injury Type`, y = Count)) +
  geom_boxplot()
```

## How have injuries rates changed over time?
### Line charts by sport nat' injury estimates
**Shows DECREASE in injuries**
```{r}
national_time <- natl_injury_estimates %>% 
  pivot_longer(cols = -X1, names_to = "Year", values_to = "Injury") %>% 
  rename(Sport = X1) %>% 
  filter(Sport != "Overall total")

temp <- national_time$Year
  
temp <- temp %>% 
  str_replace_all(pattern = "[[:punct:]]\\d{2}$", replacement = "") %>% 
  as.integer()

national_time <- national_time %>% 
  select(-Year) %>% 
  bind_cols(temp) %>% 
  rename(Year = `...3`)
```

```{r}
national_time %>% 
  ggplot(aes(x = Year, y = Injury, color = Sport)) +
  geom_line()
```

```{r}
national_time %>% 
  filter(Injury > 444281) %>% 
  ggplot(aes(x = Year, y = Injury, color = Sport)) +
  geom_line()
```


### Scatterplot diagnosis by year, color by type
```{r}
diagnosis_by_year <- diagnosis_by_year %>% 
  rename(Diagnosis = X1) %>% 
  filter(Diagnosis != "Total") %>% 
  pivot_longer(cols = !Diagnosis, names_to = "Year", values_to = "Percent")
```

```{r}
temp_year <- diagnosis_by_year$Year

temp_perc <- diagnosis_by_year$Percent
  
temp_year <- temp_year %>% 
  str_replace_all(pattern = ",", replacement = "") %>% 
  str_replace_all(pattern = "[[:punct:]]\\d{2}", replacement = "") %>% 
  as.integer()

temp_perc <- temp_perc %>% 
  str_replace_all(pattern = "%", replacement = "") %>% 
  as.numeric()

diagnosis_by_year <- diagnosis_by_year %>% 
  select(-Year, -Percent) %>% 
  bind_cols(temp_year, temp_perc) %>% 
  rename(Year = `...2`) %>% 
  rename(Percent = `...3`)
```

```{r}
diagnosis_by_year %>% 
  ggplot(aes(x = Year, y = Percent, color = Diagnosis)) +
  facet_wrap(~ Diagnosis) +
  geom_point() +
  geom_smooth()
```


# Participation
## Who participates in sports?
Choropleth map of participation
Bar chart of participation by gender

## How many people participate in sports?
Line graph of sports

# Conclusion
## Injuries and participation over time
## Stacked Bar Graph: Public opinion
```{r}
text_statements <- c("Players who suffer a head injury should be required to take a minimum set amount of time off from playing to recover.",
                     "Helmets should be changed to better protect against concussions.",
                     "The risks of playing football are widely known and players participate at their own risk.",
                     "There should be a standardized test used to determine if and when injured players at all levels may return to the field.",
                     "Aggressive tackles which are more prone to leading to head injuries should be restricted in youth football.")

survey_levels <- c(
  "Strongly agree",
  "Somewhat agree",
  "Somewhat disagree",
  "Strongly disagree",
  "Not at all sure"
)
```

```{r}
# Getting rid of Statement column to convert numbers to percents
survey_interim <- survey_tbi_effects %>% 
  select(-Statement)

# Calculating percentages
data_percentage <- survey_interim %>% 
  apply(2,
        function(x){
          perc <- x / sum(x)
          round(perc, digits = 2)
          })

# Converting to tibble and adding final column so it equals 1
diff_interim <- data_percentage %>% 
  as_tibble() %>% 
  select_if(is.numeric) %>% 
  mutate(sum = rowSums(.[1:5])) %>% 
  mutate(diff = 1 - sum)

# Saving into data_percentage
data_percentage <- diff_interim %>% 
  select(-sum)
  

# Adding the statement column back
data_percentage <- cbind(survey_tbi_effects$Statement, data_percentage) %>% 
  as_tibble()  %>% 
  slice(1:5)


# Renaming column
data_percentage <- data_percentage %>% 
  rename(Statement = `survey_tbi_effects$Statement`)

# Pivot long
long_percentage <- data_percentage %>% 
  pivot_longer(cols = contains("r")) %>% 
  rename(response = name)

```

```{r}
set_colors <- c("#91b86e", "#956eb8", "#b86e90", "#6eb895", "#b8816e", "#b6b86e")
xlabels <- c("1", "2", "3", "4", "5")

# Making graph
stacked_survey <- long_percentage %>% 
  ggplot(aes(x = Statement, y = value, fill = response)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = set_colors, name = "Response") +
  scale_x_discrete(labels = xlabels) +
  labs(title = "Survey Responses on The Long-Term Effects of Concussions in Sports",
       subtitle = "2014",
       caption = "Source: Harris Interactive | The Harris Poll #92",
       y = "") +
  theme_light()

stacked_survey
```
